Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2015-01-28 draft_lm1.tex
fixed table formatting
Some small tweaks; #9 still pending; will do later today

\subsection{The dataset} 
The data used in this paper consists of  Dengue weekly notified cases time 
series for the city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical symptoms. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013). The incidence is normalized for the parameter estimation 
procedures. The normalizatio was done by dividing the number os cases reported 
by the total city population at each year as given by the census.

@@

We explore the approach of Ederer and Mantel~\cite{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\cite{mantel} show that one can construct a $100\alpha 
\%$ confidence interval for \rr by noting that
\begin{align}
\label{eq:confRt}
 R_t^{(L)} &= \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})}\nonumber \\
 R_t^{(U)} &= \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}
\end{align}
Many authors have derived $c_{\alpha}(\theta_t)$ following orthodox approaches  
(see for example~\cite{wilson} and~\cite{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 - 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior distributions
for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in Text S1.

@@

As a side note, the Bayesian approach presented in this 
paper will give similar results to those of 
\cite{wilson} and \cite{wilson} for $Y_{t+1}$ and $Y_t >> 1$.
Under the flat  uniform prior for $\theta_t$, the Bayesian posterior 
credibility 
interval is nearly indistinguishable from the confidence interval proposed by 
\cite{clopper} for $Y_{t+1}, Y_t > 20$.
Note that the $Beta(1, 1)$ uniform prior for $\theta_t$ constitutes a poor 
prior 
choice mainly because the induced distribution for \rr is only well-defined for 
$b_0 > 2$.

@@

\begin{table}[!ht]
\caption{
\bf{Median Attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. $^\dag$: Year corresponds to the start of the epidemic. 
Sometimes the peak of cases occur in the following year.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & median Attack Ratio & $S_0$ \\
\hline
1996 & 38.64\% (17.18\%-54.19\%) & 0.171\%\\
1997 & 87.31\% (73.83\%-87.31\%) & 0.273\%\\
1998 & 49.86\% (49.19\%-49.86\%) & 0.142\%\\
1999 & 10.76\% (3.72\%-20.24\%) & 0.345\%\\
2000 & 25.49\% (24.47\%-26.98\%) & 1.547\%\\
2001 & 48.12\% (46.57\%-49.15\%) & 4.951\%\\
2005 & 14.74\% (10.23\%-21.38\%) & 1.472\%\\
2006 & 11.22\% (8.00\%-14.41\%) & 2.810\%\\
2007 & 15.08\% (14.76\%-15.38\%) & 13.451\%\\
2008 & 13.73\% (3.10\%-30.84\%) & 0.672\%\\
2010 & 18.30\% (17.34\%-19.33\%) & 4.540\%\\
2011 & 8.59\% (8.20\%-9.36\%) & 21.482\%\\
2012 & 13.92\% (12.71\%-14.92\%) & 6.208\%\\

@@

\section*{Supporting Information Legends}
%
% Please enter your Supporting Information captions below in the following format:
%\item{\bf Figure SX. Enter mandatory title here.} Enter optional descriptive information here.
% 
%\begin{description}
%\item {\bf}
%\item {\bf}
%\end{description}

>>>>

\subsection*{The data set} 
The data used in this paper consists of  Dengue weekly notified cases time 
series for the city of Rio de Janeiro from 1996 to 2014. The cases are notified 
based only on clinical symptoms. Laboratory confirmation and serotype 
information are available only for a very small sample and only on recent 
years (2010-2013). The incidence is normalized for the parameter estimation 
procedures. The normalization was done by dividing the number os cases reported 
by the total city population at each year as given by the census.

@@

We explore the approach of Ederer and Mantel~\cite{mantel}, whose objective is 
to obtain 
confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ 
and define $S = Y_{t} + Y_{t+1}$.
The authors note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that 
$Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that 
$Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
Ederer and Mantel (1974)~\cite{mantel} show that one can construct a $100\alpha 
\%$ confidence interval for \rr by noting that
\begin{equation}
\label{eq:confRt}
 R_t^{(L)} = \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})} \quad \text{and} \quad R_t^{(U)} = \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}\nonumber \\
\end{equation}

Many authors have derived $c_{\alpha}(\theta_t)$ following orthodox approaches  
(see for example~\cite{wilson} and~\cite{clopper}) mainly for simplicity.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ 
posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the 
binomial likelihood in (\ref{eq:binlike}), the posterior distribution for 
$\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Combining equations~(\ref{eq:confRt}) and~(\ref{eq:thetapost}) 
tells us that the induced posterior distribution of $R_t$ is 
a beta prime (or inverted beta) with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 
=  Y_t + b_0$.
The density of the induced distribution is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 - 
1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr is $a_1/(b_1 - 1)$ and its variance is 
$a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.
Sampling from the posterior in~(\ref{eq:densityMantel}) can be made 
straightforward by first sampling from~(\ref{eq:thetapost}) and then applying 
the transform in~(\ref{eq:confRt}).
Also, one can choose $a_0$ and $b_0$ so as to elicit meaningful prior distributions
for $R_t$.
We show how to elicit the prior for $R_t$ from specified prior mean and variance
or coefficient of variation in Text S1.

@@

As a side note, the Bayesian approach presented in this 
paper will give similar results to those of 
\cite{wilson} and \cite{clopper} for $Y_{t+1}$ and $Y_t >> 1$.
Under the flat  uniform prior for $\theta_t$, the Bayesian posterior 
credibility 
interval is nearly indistinguishable from the confidence interval proposed by 
\cite{clopper} for $Y_{t+1}, Y_t > 20$.
Note that the $Beta(1, 1)$ uniform prior for $\theta_t$ constitutes a poor 
prior 
choice mainly because the induced distribution for \rr is only well-defined for 
$b_0 > 2$.

@@

\begin{table}[!ht]
\caption{
{\bf Median Attack ratio and 95\% credibility intervals calculated according to 
(\ref{eq:AR2})}. Values are presented as percentage of total population. 
$^\dag$: Year corresponds to the start of the epidemic. 
Sometimes the peak of cases occur in the following year.
$^\ddag$: Susceptible fraction.}
\begin{center}
\begin{tabular}{c|c|c}
\hline
Year$^\dag$ & Median attack ratio (\%) & $S_0^\ddag$ (\%)\\
\hline
1996 & 38.64 (17.18-54.19) & 0.171\\
1997 & 87.31 (73.83-87.31) & 0.273\\
1998 & 49.86 (49.19-49.86) & 0.142\\
1999 & 10.76 (3.72-20.24) & 0.345\\
2000 & 25.49 (24.47-26.98) & 1.547\\
2001 & 48.12 (46.57-49.15) & 4.951\\
2005 & 14.74 (10.23-21.38) & 1.472\\
2006 & 11.22 (8.00-14.41) & 2.810\\
2007 & 15.08 (14.76-15.38) & 13.451\\
2008 & 13.73 (3.10-30.84) & 0.672\\
2010 & 18.30 (17.34-19.33) & 4.540\\
2011 & 8.59 (8.20-9.36) & 21.482\\
2012 & 13.92 (12.71-14.92) & 6.208\\

@@

%\section*{Supporting Information Legends}
%
% Please enter your Supporting Information captions below in the following format:
%\item{\bf Figure SX. Enter mandatory title here.} Enter optional descriptive information here.
% 
%\begin{description}
%\item {\bf}
%\item {\bf}
%\end{description}
